---
title: "PA1_template.Rmd"
author: "me"
date: "9/27/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, the data and packages used to process it must be loaded

```{r, message = FALSE, warning = FALSE}
data <- read.csv("activity.csv")
library(lubridate)
library(plyr)
library(dplyr)
library(tidyverse)
library(ggplot2)
```

The data is processed using dplyr to calculate the mean, median, and sum of the steps taken each day. A data table called "stepsbyday" is created, which holds this information in variables "stepsbyday", "medsteps",  and "sumsteps". A second data table called "stepsbydayf" is created for data with the missing values filled in (the rule used will be covered below). 

```{r}
#calculating mean, medium, and sum of steps by day
stepsbyday <- data%>%
      select(date, steps)%>%
      drop_na()%>%
      group_by(date)%>%
      summarize(avgsteps = mean(steps))
medsteps <- data%>%
      select(date, steps)%>%
      drop_na()%>%
      group_by(date)%>%
      summarize(medsteps = median(steps))%>%
      select(medsteps)
sumsteps <- data%>%
      select(date, steps)%>%
      drop_na()%>%
      group_by(date)%>%
      summarize(sumsteps = sum(steps))%>%
      select(sumsteps)
stepsbyday <- cbind(stepsbyday, medsteps, sumsteps)
#calculating mean steps across each time interval
across <- data%>%
      select(steps, interval)%>%
      drop_na()%>%
      group_by(interval)%>%
      summarize(meansteps = mean(steps))
across <- as.data.frame(across)
names(across) <- c("interval", "steps")
#replacing missing values
missing <- data[is.na(data$steps),]
missing <- missing %>%
      arrange(interval)%>%
      select(steps, interval)
added <- join(missing, across, by = "interval", match = "first")
names(added) <- c("interval", "removeme", "steps")
added <- added%>%
      select(interval, steps)
addtome <- data[is.na(data$steps),]
addtome <- addtome%>%
      arrange(interval)%>%
      select(date, interval)
alldone <- cbind(addtome, added$steps)
data_fixed <- left_join(data, alldone, by = c("date", "interval"))
data_fixed[is.na(data_fixed)] = 0
names(data_fixed) = c("steps1", "date", "interval", "steps2")
dataf <- data_fixed%>%
      mutate(stepsf = (steps1 + steps2))
dataf <- dataf%>%
   select(stepsf, interval, date)
dataf$date <- ymd(data$date)
#calculating mean, median, sum steps of filled in data
stepsbydayf <- dataf%>%
   select(date, stepsf)%>%
   group_by(date)%>%
   summarize(avgsteps = mean(stepsf))
medstepsf <- dataf%>%
   select(date, stepsf)%>%
   group_by(date)%>%
   summarize(medsteps = median(stepsf))%>%
   select(medsteps)
sumstepsf <- dataf%>%
   select(date, stepsf)%>%
   group_by(date)%>%
   summarize(sumstepsf = sum(stepsf))%>%
   select(sumstepsf)
stepsbydayf <- cbind(stepsbydayf, medstepsf, sumstepsf)
```

GGplot2 is now used to create a histogram displaying the sum steps by day
The mean and median steps per day are also displayed.

```{r fig.path = "figures/"}
ggplot(stepsbyday)+
      geom_histogram(color = "black", fill = "grey", aes(x = sumsteps))
```
```{r}
allmean <- mean(stepsbyday$sumsteps)
allmedian <- median(stepsbyday$sumsteps)
cbind(allmean, allmedian)
```

A call to the base plotting system creates a line graph of mean steps by time interval across the entire sample.

```{r fig.path = "figures/"}
with(across, plot(interval, steps, type = "l"))
```

``` {r}
across[which.max(across$steps),]
```

Missing values were filled in using the mean steps taken across all subjects during that time interval (using only real values).
Similar steps as were taken with the original data frame "data" are taken with dataf to generate a histogram

```{r fig.path = "figures/"}
ggplot(stepsbydayf)+
   geom_histogram(color = "black", fill = "grey", aes(x = sumstepsf))
```

``` {r}
allmeanf <- mean(stepsbydayf$sumstepsf)
allmedianf <- median(stepsbydayf$sumstepsf)
cbind(allmean, allmedian)
```

We see that filling in the data leave the median and mean remarkably almost unchanged.
dplyr and grepl are used to add another variable to dataf denoting whether the day in question is a weekday or weekend

```{r}
dataf <- dataf%>%
   mutate(day = weekdays(date))%>%
   mutate(daytype = ifelse(grepl("Saturday", day)|(grepl("Sunday", day)), "Weekend", "Weekday"))
dataf$daytpe <- as.factor(dataf$daytype)
daytypemean <- dataf%>%
   select(stepsf, interval, daytype)%>%
   group_by(daytype, interval)%>%
   summarize(meansteps = mean(stepsf))
```

Finally, a faceted call to ggplot2 creates side by side line graphs comparing the mean steps by interval between weekdays and weekends.

```{r fig.path = "figures/"}
ggplot(daytypemean, aes(x = interval, y = meansteps))+
   geom_line()+
   facet_wrap(~ daytype)
```

